Estimating demographic bias on tests of children's early vocabulary

### 2022-09-08 paper/manuscript.Rmd
handle warnings, figures in text

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

@@

We focus here on the MacArthur-Bates Communicative Development Inventories (CDIs) [@fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.[^1]
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

@@

```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "t", fig.align='center', fig.width=6.5, fig.height=3, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs. age by demographic group, from the baseline Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=10) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors) +
  theme(legend.position = "top")

@@

```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right). Words at the top are more well-known by one group (e.g., females, for sex), while those at the bottom are more known by the other group (e.g., males), and the selection in the middle show roughly the median difficulty difference for those groups."}
knitr::include_graphics("figs/smGLIMMER_combo.pdf")
```

@@

```{r pruned-coefs-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
load(here("data/pruned_demo_coefs.Rdata"))
# horizontal version
pdat <- prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
  pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
                           names_to=c("type", "demo"), names_sep="_") %>%
  pivot_wider(names_from=type, values_from=value) %>%
  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo)))

@@

```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic. The 19 items in bold are those showing a difference $>3$ SD."}
knitr::include_graphics("figs/extrema2p25sd_3sd_bold.pdf")
```

>>>>

floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no

@@

We focus here on the MacArthur-Bates Communicative Development Inventories (CDIs) [@fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.[^1]
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created (Wordbank) [@frank2017; @frank2021].

@@

```{r baseline-ability-vs-age-hor, warning=F, out.width="\\linewidth", fig.align='center', fig.width=6.5, fig.height=3, fig.cap = "Language ability vs. age by demographic group, from the baseline Rasch model."}
# fig.env = "figure*", fig.pos = "t", set.cap.width=T, num.cols.cap=2,
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=10) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors) +
  theme(legend.position = "top")

@@

```{r glimmer-combo, out.width="\\linewidth", fig.width=6.5, fig.align = "center", fig.cap = "GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right). Words at the top are more well-known by one group (e.g., females, for sex), while those at the bottom are more known by the other group (e.g., males), and the selection in the middle show roughly the median difficulty difference for those groups."}
knitr::include_graphics("figs/smGLIMMER_combo.pdf")
```

@@

```{r pruned-coefs-hor, out.width="\\linewidth", fig.width=6.5, fig.height=3, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
load(here("data/pruned_demo_coefs.Rdata"))
# horizontal version
pdat <- prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
  pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
                           names_to=c("type", "demo"), names_sep="_") %>%
  pivot_wider(names_from=type, values_from=value) %>%
  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo)))

@@

```{r biased-words, include=T, out.width=".7\\linewidth", fig.align='center', fig.width=3.5, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic. The 19 items in bold are those showing a difference $>3$ SD."}
knitr::include_graphics("figs/extrema2p25sd_3sd_bold.pdf")
```
